---
title: "Plankton PDFs transformations"
output: html_notebook
---

Using a R Markdown Notebook to explore the transformation of plankton probability density functions (PDFs) to two linearly conserved variables (Biomass and total organism count)

Sample data: Luo et al. unpublished. Phytoplankton and zooplankton individual lengths from an in-situ imaging system. Data subsetted from original.

```{r}
# set working directory
setwd("/Users/jluo/Documents/NCAR/models/plankton-pdf")

d <- read.table("data/phyto_zoo_lengths_subset.txt", sep = "", header=TRUE)

d$type <- ifelse(d$type==1, "phyto", "zoo")
d$len_mm <- d$len_px / 15.1704

print(str(d))
print(head(d))
```

```{r}
plot(density(d$len_mm), main = "PDF of plankton size distributions", xlab="Organism lengths (mm), N = 302585")
```

```{r}
plot(density(log10(d$len_mm)))
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).
