---
title: "Plankton PDFs transformations"
output: html_notebook
---

Using a R Markdown Notebook to explore the transformation of plankton probability density functions (PDFs) to two linearly conserved variables (Biomass and total organism count)

Sample data: Luo et al. unpublished. Phytoplankton and zooplankton individual lengths from an in-situ imaging system. Data subsetted from original.

```{r}
# libraries
library(ggplot2)
library(plyr)

# set working directory
# setwd("/Users/jluo/Documents/NCAR/models/plankton-pdf")

# functions for geometric mean and sd
gm_mean = function(x, na.rm=TRUE, na.exclude=FALSE){
  if(na.exclude){
    x <- x[which(!is.na(x))]
    x <- x[x>0]
    res <- exp(mean(log(x), na.rm=na.rm))
    return(res)
  } else {
    x <- x[x>0]
    res <- exp(mean(log(x), na.rm=na.rm))
    return(res)
  }
}

gm_sd <- function(x, na.rm = TRUE, ...)
{
  exp(sd(log(x, ...), na.rm = na.rm, ...))
}


```

```{r}
d <- read.table("data/phyto_zoo_lengths_subset.txt", sep = "", header=TRUE)
d <- d[which(d$len_px > 0),]

d$type <- ifelse(d$type==1, "phyto", "zoo")
d$len_mm <- d$len_px / 15.1704

print(str(d))
print(head(d, 3))
```


```{r}
plot(density(d$len_mm), main = "PDF of plankton size distributions", xlab="Organism lengths (mm), N = 302585")
```

Organism lengths are then converted to biomass according to a general allometric relationship from Rodriguez and Mullin (1986):
log Biomass = 2.23 log ESD - 5.58

```{r}
len_ug <- d$len_mm * 1000
B_ug <- 2.23 * log10(len_ug) - 5.58 

d$biomass_ugC <- B_ug #/ 1000

# lengths under 0.318 mm result in a negative biomass, exclude those from this analysis
d <- d[d$len_mm >= 0.318,]

plot(density(d$biomass_ugC))
```

Convert biomass to a normalized biomass spectrum. Patt and Denman (1977) proposed the "normalized biomass spectrum (NBS)" in which biomass concentration per size interval is divided by the linear width of the interval and displayed on a logarithmic ordinate.

Conversion using equal bins on a log-scale:
```{r}
log_biomass <- sort(unique(log10(d$biomass_ugC)))
# add the last element to ensure the breaks do not get cut off
log_biomass <- c(log_biomass, round(max(log_biomass), 2) + 0.01) 


# create the levels
log_cuts = data.frame(levels=levels(cut(log_biomass, breaks = 100)))


# parse the levels into min and max
# replace every punctuation mark except [,.-] by an empty string
min_max <- unlist(strsplit(gsub("(?![,.-])[[:punct:]]", "", 
           as.character(log_cuts$levels), perl=TRUE), ","))
log_cuts$log_min <- as.numeric(min_max[seq(1, length(min_max), by=2)])
log_cuts$log_max <- as.numeric(min_max[seq(2, length(min_max), by=2)])

log_cuts$min <- 10^(log_cuts$log_min)
log_cuts$max <- 10^(log_cuts$log_max)
```

```{r}
print(head(log_cuts, 3))
print(tail(log_cuts, 3))
```
 
```{r}
breaks <- c(log_cuts$min[1], log_cuts$max)

d$bin <- cut(d$biomass_ugC, breaks=breaks) # factor

# calculate the linear length of each bin
min_max <- unlist(strsplit(gsub("(?![,.-])[[:punct:]]", "", 
           as.character(levels(d$bin)), perl=TRUE), ","))
min <- as.numeric(min_max[seq(1, length(min_max), by=2)])
max <- as.numeric(min_max[seq(2, length(min_max), by=2)])

bin_lengths <- max - min

ggplot(data=data.frame(bin_lengths=bin_lengths, n=1:length(bin_lengths))) + geom_line(aes(y=bin_lengths, x=n)) + labs(title="Bin Lengths")
```

```{r}
# assign bin lengths to bins
d$bin_lengths <- bin_lengths[as.numeric(d$bin)]

# collect bins together, calculate a mean biomass per bin and divide by the width of the interval
dd <- ddply(d, ~bin+type, function(x){
  len_px <- gm_mean(x$len_px)
  len_mm <- gm_mean(x$len_px)
  biomass_ugC <- gm_mean(x$biomass_ugC)
  n <- nrow(x)
  
  return(data.frame(n, len_px, len_mm, biomass_ugC, bin_length=unique(x$bin_length)))
})

head(dd)
```

Plot of normalized biomass. Not sure how to get this to look linear.
```{r}
dd$norm_biomass <- dd$biomass_ugC / dd$bin_length
plot(density(dd$norm_biomass))
```

```{r}
plot(density(log10(dd$norm_biomass)))
```

Notes on R Notebooks
--

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).
